Deep Learning and Bayesian Modeling

### 2021-05-24 DeepLearningAndBayesianSoftware.Rmd
added in emcee hammer, rejected inclusion in the end, too specialized

For the purposes of this document we take Bayesian software to be Stan (@standev) and PyMC3 (@salvatier2016probabilistic) with ecosystem components included that are likely to be cited. That includes the simplified syntax interfaces to Stan, brms (@brms), RstanArm (@rstanarm), and interface packages to Stan, RStan (@rstan) and PyStan (@pystan)[^1]. The analysis applies only to packages that are in current development so the venerated, and very high impact, packages like BUGS (@lunn2013bugs) and JAGS (@plummer2003jags) are not considered although including them would double our counts. Also, the ecosystems are actually much larger but they are unlikely to be cited in the research literature and we have to stop some place. For deep learning we take TensorFlow, its interface Keras (@keras) and PyTorch (@pytorch) as roughly equivalent entities for the comparison. Theano (@theano) is no longer in development. It should be noted that both PyTorch and TensorFlow have implemented HMC with NUTS for Bayesian inference but those are recent developments that have not made much of an impact yet.

@@

if (USE_CACHE) {
  library(redux)
  redis <- redux::hiredis()
  get_results <- function(url) {
    cache <- redis$GET(url) # check redis first
    if (!is.null(cache)) {
      result <- unserialize(cache)
      if (result$status_code == 200) {
        if (REPORT_PROGRESS) {
          cat(paste("\nhitting redis for", url))
        }
        return(result)
      }
    }
    random_wait <- abs(rnorm(1, 1, 1))
    if (REPORT_PROGRESS) {
      cat(paste("\ncache miss, querying:", url, "\n"))
      cat(paste(
        "\nWaiting",
        random_wait,
        "seconds to be nice to webserver\n"
      ))
    }
    Sys.sleep(random_wait)
    result <- GET(
      url,
      add_headers('X-ELS-APIKey' = API_KEY,
                  'X-ELS-Insttoken' = INSTITUTION_TOKEN)
    )
    redis$SET(url, serialize(result, NULL))
    return(result)
  }
} else { #not using cache, hacky but works with R
  get_results <- function(url) {
    random_wait <- abs(rnorm(1, 1, 1))
    if (REPORT_PROGRESS) {
      cat(paste("\nno cache used, querying:", url, "\n"))
      cat(paste(
        "\nWaiting",
        random_wait,
        "seconds to be nice to webserver\n"
      ))
    }
    Sys.sleep(random_wait)
    result <- GET(
      url,
      add_headers('X-ELS-APIKey' = API_KEY,
                  'X-ELS-Insttoken' = INSTITUTION_TOKEN)
    )
    return(result)
  }
}

@@

stan_q <- '(gelman+AND+hoffman+AND+stan)+OR+mc-stan.org'
pymc_q <- 'PyMC3+OR+(PyMC*+AND+fonnesbeck)'
rstan_q <- '(rstan+AND+NOT+mit)'
brms_q <-  '(brms+AND+burkner)'

#Display names followed by queries
pkg_query <- c('Stan', 'brms','PyStan', 'RStanArm', 'RStan','PyMC',  'TensorFlow', 'PyTorch', 'Keras',
               stan_q, brms_q, 'pystan', 'rstanarm', rstan_q, pymc_q, 'tensorflow', 'pytorch', 'Keras')
pkg_query_m <- matrix(data = pkg_query, ncol =2)
queryBayVsDeepM <- matrix(ncol = 2, nrow = 2)
queryBayVsDeepM[1,] <- c('Bayesian', paste(pkg_query_m[1:6,2], collapse = "+OR+"))
queryBayVsDeepM[2,] <- c('Deep Learning', paste(pkg_query_m[7:9,2], collapse = "+OR+"))

@@

years <- year_start:year_end
scopus.df <- data.frame(years)
for (i in 1:nrow(queryBayVsDeepM)) {
  package_name = queryBayVsDeepM[i, 1]
  query = queryBayVsDeepM[i, 2]
  if (query == '') {
    query = package_name
  }
  for (suffix in c('','_no_comp_sci')) {
    year_counts = c()
        for (year in year_start:year_end) {
      url <- paste(BASE_URL, '?query=','REF(', query,')',
                   "+AND+PUBYEAR+=+", year,
                   sep = '')
      if (suffix == '_no_comp_sci') {
        url = paste(url, '+AND+NOT+SUBJAREA(COMP)', sep = '')
      }
      result <- get_results(url)
      json_txt <- rawToChar(as.raw(strtoi(result$content, 16L)))
      data <- jsonlite::fromJSON(json_txt)
      year_counts = c(
        year_counts,
        as.numeric(data$`search-results`$`opensearch:totalResults`))
    }
    scopus.df[paste(
      package_name,
      suffix,
      sep = '')] <- year_counts
  }
}

@@

print(plot2)
```

@@

getAllSubjectCategories <- function (queryM) {
  year_start = 2006
  year_end = 2021
  years <- year_start:year_end
  accumDf = data.frame(matrix(ncol = 2, nrow = nrow(queryM) * length(years)))
  colnames(accumDf) = c('package', 'years')
  counter = 1
  article_count = 0
  article_subj_count = 0
  for (i in 1:nrow(queryM)) {
    package_name = queryM[i, 1]
    query = queryM[i, 2]
    if (!is.na(str_match(query,"^\\(")) && !is.na(str_match(query,"\\)$"))) {
      query = str_match(query, "^\\((.*)\\)$")[1,2] #remove extra brackets, breaks scopus
    }
    accumDf[counter:((counter - 1) + length(years)),]$years <- years
    accumDf[counter:((counter - 1) + length(years)),]$package <- 
      rep(package_name, length(years))
    for (year in year_start:year_end) {
      url <-
        paste(
          BASE_URL,
          '?query=',
          'REF(',
          query,
          ')',
          "+AND+PUBYEAR+=+",
          year,
          '&facets=subjarea(count=101)',
          sep = ''
        )
      result <- get_results(url)
      if (result$status_code != 200) {
        print(sprintf("got non 200 status from query: %d", result$status_code))
        stop()
      }
      json_txt <- rawToChar(as.raw(strtoi(result$content, 16L)))
      data <- jsonlite::fromJSON(json_txt)
      article_count <-
        as.numeric(data$`search-results`$`opensearch:totalResults`) + article_count
      facet_count <- length(data$`search-results`$facet$category$name)
      j <- 1
      while (j < facet_count) {
        name <- data$`search-results`$facet$category$label[j]
        name <- str_replace(name, " \\(all\\)", "")
        hitCount <- as.numeric(data$`search-results`$facet$category$hitCount[j])
        article_subj_count <- hitCount + article_subj_count
        if (!name %in% colnames(accumDf)) {
          accumDf[name] <- rep(0, length(years) * nrow(queryM))
          #print(paste("name=",name,", count=",hitCount))
        }
        accumDf[counter,][[name]] <- hitCount
        j <- j+ 1
      }
      counter = counter + 1
    }
  }
  if (REPORT_PROGRESS) {
    print(sprintf("%d paperXsubject with %d total papers for %.3f avg subjects per paper",
                  article_subj_count,
                  article_count,
                  article_subj_count/article_count))
  }
  return(accumDf)
}

>>>>

For the purposes of this document we take Bayesian software to be Stan (@standev) and PyMC3 (@salvatier2016probabilistic) with ecosystem components included that are likely to be cited. That includes the simplified syntax interfaces to Stan, brms (@brms), RstanArm (@rstanarm), and interface packages to Stan, RStan (@rstan) and PyStan (@pystan)[^1]. The analysis applies only to packages that are in current development so the venerated, and very high impact, packages like BUGS (@lunn2013bugs) and JAGS (@plummer2003jags) are not considered although including them would double our counts. There is also the best named Bayesian package of all time, "emcee: The MCMC Hammer", (@foreman2013emcee) which enjoys tremendous use in the astrophysics community but is too specialized to be considered a general Bayesian package so it is not included. Also, the ecosystems are actually much larger but they are unlikely to be cited in the research literature and we have to stop some place. For deep learning we take TensorFlow, its interface Keras (@keras) and PyTorch (@pytorch) as roughly equivalent entities for the comparison. Theano (@theano) is no longer in development. It should be noted that both PyTorch and TensorFlow have implemented HMC with NUTS for Bayesian inference but those are recent developments that have not made much of an impact yet.

@@

if (USE_CACHE) {
  library(redux)
  redis <- redux::hiredis()
  get_results <- function(url) {
    cache <- redis$GET(url) # check redis first
    if (!is.null(cache)) {
      result <- unserialize(cache)
      if (result$status_code == 200) {
        if (REPORT_PROGRESS) {
          cat(paste("\nhitting redis for", url))
        }
        return(result)
      }
    }
    random_wait <- abs(rnorm(1, 1, 1))
    if (REPORT_PROGRESS) {
      cat(paste("\ncache miss, querying:", url, "\n"))
      cat(paste(
        "\nWaiting",
        random_wait,
        "seconds to be nice to webserver\n"
      ))
    }
    Sys.sleep(random_wait)
    result <- GET(
      url,
      add_headers('X-ELS-APIKey' = API_KEY,
                  'X-ELS-Insttoken' = INSTITUTION_TOKEN)
    )
    if (result$status_code != 200) { 
      stop(sprintf("Got error from search, got status code %d",result$status_code))
    }
    redis$SET(url, serialize(result, NULL))
    return(result)
  }
} else { #not using cache, hacky but works with R
  get_results <- function(url) {
    random_wait <- abs(rnorm(1, 1, 1))
    if (REPORT_PROGRESS) {
      cat(paste("\nno cache used, querying:", url, "\n"))
      cat(paste(
        "\nWaiting",
        random_wait,
        "seconds to be nice to webserver\n"
      ))
    }
    Sys.sleep(random_wait)
    result <- GET(
      url,
      add_headers('X-ELS-APIKey' = API_KEY,
                  'X-ELS-Insttoken' = INSTITUTION_TOKEN)
    )
    if (result$status_code != 200) { 
      stop(sprintf("Got error from search, got status code %d",result$status_code))
    }
    return(result)
  }
}

@@

stan_q <- '(gelman+AND+hoffman+AND+stan)+OR+mc-stan.org'
pymc_q <- 'PyMC3+OR+(PyMC*+AND+fonnesbeck)'
rstan_q <- 'rstan+AND+NOT+mit'
brms_q <-  'brms+AND+burkner'
emcee_q <- 'emcee+AND+foreman-mackey'

#Display names followed by queries
pkg_query <- c('emcee', 'Stan', 'brms','PyStan', 'RStanArm', 'RStan','PyMC',  'TensorFlow', 'PyTorch', 'Keras',
               emcee_q, stan_q, brms_q, 'pystan', 'rstanarm', rstan_q, pymc_q, 'tensorflow', 'pytorch', 'Keras')
pkg_query_m <- matrix(data = pkg_query[c(-1,-11)], ncol =2) #emcee removed
queryBayVsDeepM <- matrix(ncol = 2, nrow = 2)
queryBayVsDeepM[1,] <- c('Bayesian', 
                         paste0('(', paste(pkg_query_m[1:6,2], collapse = ")+OR+("),')'))
queryBayVsDeepM[2,] <- c('Deep Learning', 
                         paste0('(', paste(pkg_query_m[7:9,2], collapse = ")+OR+("), ')'))

@@

years <- year_start:year_end
scopus.df <- data.frame(years)
#queryM = pkg_query_m
queryM = queryBayVsDeepM
for (i in 1:nrow(queryM)) {
  package_name = queryM[i, 1]
  query = queryM[i, 2]
  for (suffix in c('','_no_comp_sci')) {
#  for (suffix in c('')) {
    year_counts = c()
    for (year in year_start:year_end) {
      url <- paste(BASE_URL, '?query=','REF(', query,')',
                   "+AND+PUBYEAR+=+", year,
                   sep = '')
      if (suffix == '_no_comp_sci') {
        url = paste(url, '+AND+NOT+SUBJAREA(COMP)', sep = '')
      }
      result <- get_results(url)
      json_txt <- rawToChar(as.raw(strtoi(result$content, 16L)))
      data <- jsonlite::fromJSON(json_txt)
      year_counts = c(
        year_counts,
        as.numeric(data$`search-results`$`opensearch:totalResults`))
    }
    scopus.df[paste(
      package_name,
      suffix,
      sep = '')] <- year_counts
  }
}

@@

   log_plot = plot2 + 
     scale_y_continuous(breaks=c(1, 10, 100, 1000, 10000, 100000, 1e+06, 1e+07, 1e+08), 
                     trans = scales::log_trans())
print(plot2)
```

@@

getAllSubjectCategories <- function (queryM) {
  year_start = 2006
  year_end = 2021
  years <- year_start:year_end
  accumDf = data.frame(matrix(ncol = 2, nrow = nrow(queryM) * length(years)))
  colnames(accumDf) = c('package', 'years')
  counter = 1
  article_count = 0
  article_subj_count = 0
  for (i in 1:nrow(queryM)) {
    package_name = queryM[i, 1]
    query = queryM[i, 2]
    accumDf[counter:((counter - 1) + length(years)),]$years <- years
    accumDf[counter:((counter - 1) + length(years)),]$package <- 
      rep(package_name, length(years))
    for (year in year_start:year_end) {
      url <-
        paste(
          BASE_URL,
          '?query=',
          'REF(',
          query,
          ')',
          "+AND+PUBYEAR+=+",
          year,
          '&facets=subjarea(count=101)',
          sep = ''
        )
      result <- get_results(url)
      json_txt <- rawToChar(as.raw(strtoi(result$content, 16L)))
      data <- jsonlite::fromJSON(json_txt)
      article_count <-
        as.numeric(data$`search-results`$`opensearch:totalResults`) + article_count
      facet_count <- length(data$`search-results`$facet$category$name)
      j <- 1
      while (j < facet_count) {
        name <- data$`search-results`$facet$category$label[j]
        name <- str_replace(name, " \\(all\\)", "")
        hitCount <- as.numeric(data$`search-results`$facet$category$hitCount[j])
        article_subj_count <- hitCount + article_subj_count
        if (!name %in% colnames(accumDf)) {
          accumDf[name] <- rep(0, length(years) * nrow(queryM))
          #print(paste("name=",name,", count=",hitCount))
        }
        accumDf[counter,][[name]] <- hitCount
        j <- j+ 1
      }
      counter = counter + 1
    }
  }
  if (REPORT_PROGRESS) {
    print(sprintf("%d paperXsubject with %d total papers for %.3f avg subjects per paper",
                  article_subj_count,
                  article_count,
                  article_subj_count/article_count))
  }
  return(accumDf)
}
